Charity During Crackdown: Analyzing the Impact of State Repression of NGOs on Philanthropy

### 2020-01-24 manuscript/manuscript.Rmd
Final edits pre-submission

---
title: "Charity During Crackdown: Analyzing the Impact of State Repression of NGOs on Philanthropy"
short-title: Charity During Crackdown
author:
- name: Suparna Chaudhry
  affiliation: Christopher Newport University
  email: suparna.chaudhry@cnu.edu
  url: http://www.suparnachaudhry.com/
  twitter: SuparnaChaudhry
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
  url: https://www.andrewheiss.com/
  twitter: andrewheiss
date: "`r format(Sys.time(), '%B %e, %Y')`"
git-repo: https://github.com/andrewheiss/ngo-crackdowns-philanthropy
reference-section-title: References
toc: false
title-page: true
abstract: >-
  Government restrictions on NGOs have become increasingly pervasive across the globe. With reduced official aid and foundation funding available for NGOs, and an absence of philanthropic culture in many countries across the Global South, various NGOs are seeking funds from individual private donors in Western countries. How do legal restrictions on NGOs impact patterns of private philanthropic giving and individual donor decision-making? While existing research has shown that donors rely on various heuristics from social networks and organization-level attributes, we argue that domestic political environments of NGO host countries can also serve as an important structural-level heuristic that can influence donor behavior. Using a survey experiment, we find that while crackdowns do not directly influence the likelihood of donation on their own, donors are willing to donate substantially more to legally besieged NGOs abroad. We discuss the implications of our results on the sustainability of organizations working in repressive countries.
thanks: >-
  We thank Mark Buntaine, Michael DeCrescenzo, Jeffrey Friedman, Thomas Leeper, and Tristan Mahr for their helpful comments and input. We also thank Samantha Camilletti and Joshua Dutro for their excellent research assistance. This experiment received approval by the human subjects research committees at both Christopher Newport University (018.042) and Brigham Young University (E18104). We preregistered our hypotheses and research design at the Open Science Framework, and our preregistration protocol is available at <https://osf.io/dx973/>.
author-note: >-
  Suparna Chaudhry, PhD, Assistant Professor of Political Science, Christopher Newport University; Andrew Heiss, PhD, Assistant Professor of Public Management, Andrew Young School of Policy Studies, Georgia State University. 
additional-info: Figures should appear in color online only, not in print.
correspondence: >-
  Correspondence concerning this article should be addressed to Suparna Chaudhry, PhD, Department of Political Science, Christopher Newport University, Newport News, VA 23606, 757-594-7821. Contact: suparna.chaudhry@cnu.edu
word-count: "4,994"
keywords: international NGOs, donor behavior, philanthropy, human rights, repression
---

@@

@fig:likely-diffs(B–C) shows the difference in the likelihood of donation for both crackdown conditions across NGO issue areas (H~2a~) and funding sources (H~3a~). Crackdown has almost no effect on the likelihood of donating to human rights NGOs (`r filter(likely, title == "Human rights")$p_neq0`). We hypothesized that the positive crackdown effect would be stronger for humanitarian NGOs, and while crackdown does have a positive effect under the humanitarian frame, the probability this difference is greater than zero is below our 95% threshold (`r filter(likely, title == "Humanitarian assistance")$p_neq0`). We do find some evidence for H~3a~, however. Crackdown has little effect on the likelihood of individual preferences to donate to NGOs receiving government funding (`r filter(likely, title == "Government")$p_neq0`), but it does increase the likelihood of donating to privately-funded NGOs by `r percent_0(filter(likely, title == "Private")$pct_change)` (`r filter(likely, title == "Private")$delta`; `r filter(likely, title == "Private")$p_neq0`).

While the crackdown frame has no effect on the probability of donating to human rights NGOs alone, conditioning this finding on the source of NGO funding reveals competing trends. @fig:likely-diffs(D) shows the difference in donation likelihood across all experimental conditions. When donors know that a human rights organization is primarily government-funded, they are more likely to donate in the *absence* of crackdowns than when the NGO faces legal difficulties (`r filter(likely, title == "Human rights | Government")$delta`; `r filter(likely, title == "Human rights | Government")$p_neq0`). In contrast, when donors know that an organization is privately-funded, they are more likely to donate when there *is* a legal crackdown than when the NGO faces no legal hurdles (`r filter(likely, title == "Human rights | Private")$delta`; `r filter(likely, title == "Human rights | Private")$p_neq0`). For humanitarian assistance NGOs, however, legal crackdowns do not have a significant effect on the likelihood of donation when nesting funding and issue frames.

These results give partial support to H~3a~. The source of funding has little influence on the preference to donate to humanitarian NGOs, but on average, donors are substantially more likely to want to give to privately-funded human rights NGOs than government-funded organizations. Additionally donors appear to both punish government-funded human rights NGOs facing crackdown and rally behind privately-funded NGOs facing those same hurdles. Respondents described their reasoning for the former as "[i]f the non-profit works in countries where non-profit activities are heavily restricted there's a good amount of waste and shortfalls in their activities and ability to assist others."[^3154467] Respondents also explained their motivation for increased support to privately funded NGOs as "[t]hey [IRC] are doing good work in countries where it is tough for groups like them to operate and they need all the help they can get."[^3025921] Individual donors thus seem to be more willing to support besieged human rights organizations when they are unencumbered by government funds.

@@

Combining issue areas and funding sources provides more texture. We previously found that donors seem to (1) punish government-funded human rights NGOs and (2) reward privately-funded human rights NGOs when they face legal issues. This finding also applies to preferences regarding the amount donors donate. Respondents donated `r md_dollar(filter(amount, title == "Human rights | Private")$mu_diff)` more to besieged privately-funded human rights NGOs compared to the no crackdown control condition (`r filter(amount, title == "Human rights | Private")$p_neq0`), increasing their donations by `r percent_0(filter(amount, title == "Human rights | Private")$pct_change)`. Respondents explained that they chose to give additional money to privately-funded human rights NGOs precisely because "the country limits non-profits."[^13685] The punishment mechanism found previously does not translate to the amount donated—while there is a decrease of `r md_dollar(abs(filter(amount, title == "Human rights | Government")$mu_diff))` in average donations to government-funded human rights NGOs facing crackdown, the chance this difference is less than zero is lower than our threshold (`r filter(amount, title == "Human rights | Government")$p_neq0`).

@@

In contrast, crackdowns have no substantial effect on the amount donated to privately-funded humanitarian NGOs (`r filter(amount, title == "Humanitarian assistance | Private")$p_neq0`). No respondent in these conditions mentioned funding sources in their justification, and many explained that they would not give solely because of the crackdown and that they "would keep all \$100 because it would not even go to helping humanitarian efforts due to the new laws."[^8995923] It is possible that donors see government-funded humanitarian NGOs as more legitimate and more capable of handling difficult legal restrictions than privately-funded organizations. Donors might assume that an INGO which receives substantial money from USAID would have the support of the U.S. government when facing legal challenges, while an NGO funded by small donors would be less able to confront such challenges. Further research is needed to probe this trend.

@@

Dozens of countries have sought to restrict NGO funding and programming. Between the lack of a local philanthropic culture, an increasingly authoritarian approach that restricts space for domestic philanthropy, and the withdrawal of official governmental aid, many INGOs have turned to foreign private donors as an additional funding source. Since seeking out perfect information about a nonprofit and its activities, especially internationally, is costly and time-consuming, donors use heuristics when deciding whom to support. Existing research has shown that donors take cues and heuristics from social networks, but can other informational cues also influence donor preferences? We argue that domestic political environments of NGO host countries can indeed serve as one such heuristic. Our results show that while crackdowns do not directly influence individual preferences of donation on their own, respondents indicated willingness to donate `r percent_0(filter(amount, title == "Crackdown")$pct_change)` more to legally restricted NGOs. This effect persists when mediated by other heuristics, with respondents giving `r percent_0(filter(amount, title == "Humanitarian assistance")$pct_change)` more to humanitarian NGOs, and `r percent_0(filter(amount, title == "Human rights | Private")$pct_change)` more to privately-funded human rights NGOs. Anti-NGO crackdowns can thus serve as a signal of NGO effectiveness.

>>>>

---
title: "Charity During Crackdown: Analyzing the Impact of State Repression of NGOs on Philanthropy"
short-title: Charity During Crackdown
author:
- name: Suparna Chaudhry
  affiliation: Christopher Newport University
  email: suparna.chaudhry@cnu.edu
  url: http://www.suparnachaudhry.com/
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
  url: https://www.andrewheiss.com/
date: "`r format(Sys.time(), '%B %e, %Y')`"
git-repo: https://github.com/andrewheiss/ngo-crackdowns-philanthropy
reference-section-title: References
toc: false
title-page: true
abstract: >-
  Government restrictions on NGOs have become increasingly pervasive across the globe. With reduced official aid and foundation funding available for NGOs, and an absence of philanthropic culture in many countries across the Global South, various NGOs are seeking funds from individual private donors in Western countries. How do legal restrictions on NGOs impact patterns of private philanthropic giving and individual donor decision-making? While existing research has shown that donors rely on various heuristics from social networks and organization-level attributes, we argue that domestic political environments of NGO host countries can also serve as an important structural-level heuristic that can influence donor behavior. Using a survey experiment, we find that while crackdowns do not directly influence the likelihood of donation on their own, donors are willing to donate substantially more to legally besieged NGOs abroad. We discuss the implications of our results on the sustainability of organizations working in repressive countries.
thanks: >-
  We thank Mark Buntaine, Michael DeCrescenzo, Jeffrey Friedman, Thomas Leeper, and Tristan Mahr for their helpful comments and input. We also thank Samantha Camilletti and Joshua Dutro for their excellent research assistance. This experiment received approval by the human subjects research committees at both Christopher Newport University (018.042) and Brigham Young University (E18104). We preregistered our hypotheses and research design at the Open Science Framework, and our preregistration protocol is available at <https://osf.io/dx973/>.
author-note: >-
  Suparna Chaudhry, PhD, Assistant Professor of Political Science, Christopher Newport University; Andrew Heiss, PhD, Assistant Professor of Public Management, Andrew Young School of Policy Studies, Georgia State University. 
additional-info: Figures should appear in color online only, not in print.
correspondence: >-
  Correspondence concerning this article should be addressed to Suparna Chaudhry, PhD, Department of Political Science, Christopher Newport University, Newport News, VA 23606, 757-594-7821. Contact: suparna.chaudhry@cnu.edu
word-count: "4,999"
keywords: international NGOs, donor behavior, philanthropy, human rights, repression
---

@@

@fig:likely-diffs(B–C) shows the difference in the likelihood of donation for both crackdown conditions across NGO issue areas (H~2a~) and funding sources (H~3a~). Crackdown has almost no effect on the likelihood of donating to human rights NGOs (`r filter(likely, title == "Human rights")$p_neq0`). We hypothesized that the positive crackdown effect would be stronger for humanitarian NGOs, and while crackdown does have a positive effect under the humanitarian frame, the probability this difference is greater than zero is below our 95% threshold (`r filter(likely, title == "Humanitarian assistance")$p_neq0`). We do find some evidence for H~3a~, however. Crackdown has little effect on the likelihood of individual preferences to donate to NGOs receiving government funding (`r filter(likely, title == "Government")$p_neq0`), but it does increase the likelihood of donating to privately funded NGOs by `r percent_0(filter(likely, title == "Private")$pct_change)` (`r filter(likely, title == "Private")$delta`; `r filter(likely, title == "Private")$p_neq0`).

While the crackdown frame has no effect on the probability of donating to human rights NGOs alone, conditioning this finding on the source of NGO funding reveals competing trends. @fig:likely-diffs(D) shows the difference in donation likelihood across all experimental conditions. When donors know that a human rights organization is primarily government-funded, they are more likely to donate in the *absence* of crackdowns than when the NGO faces legal difficulties (`r filter(likely, title == "Human rights | Government")$delta`; `r filter(likely, title == "Human rights | Government")$p_neq0`). In contrast, when donors know that an organization is privately funded, they are more likely to donate when there *is* a legal crackdown than when the NGO faces no legal hurdles (`r filter(likely, title == "Human rights | Private")$delta`; `r filter(likely, title == "Human rights | Private")$p_neq0`). For humanitarian assistance NGOs, however, legal crackdowns do not have a significant effect on the likelihood of donation when nesting funding and issue frames.

These results give partial support to H~3a~. The source of funding has little influence on the preference to donate to humanitarian NGOs, but on average, donors are substantially more likely to want to give to privately funded human rights NGOs than government-funded organizations. Additionally donors appear to both punish government-funded human rights NGOs facing crackdown and rally behind privately funded NGOs facing those same hurdles. Respondents described their reasoning for the former as "[i]f the non-profit works in countries where non-profit activities are heavily restricted there's a good amount of waste and shortfalls in their activities and ability to assist others."[^3154467] Respondents also explained their motivation for increased support to privately funded NGOs as "[t]hey [IRC] are doing good work in countries where it is tough for groups like them to operate and they need all the help they can get."[^3025921] Individual donors thus seem to be more willing to support besieged human rights organizations when they are unencumbered by government funds.

@@

Combining issue areas and funding sources provides more texture. We previously found that donors seem to (1) punish government-funded human rights NGOs and (2) reward privately funded human rights NGOs when they face legal issues. This finding also applies to preferences regarding the amount donors donate. Respondents donated `r md_dollar(filter(amount, title == "Human rights | Private")$mu_diff)` more to besieged privately funded human rights NGOs compared to the no crackdown control condition (`r filter(amount, title == "Human rights | Private")$p_neq0`), increasing their donations by `r percent_0(filter(amount, title == "Human rights | Private")$pct_change)`. Respondents explained that they chose to give additional money to privately funded human rights NGOs precisely because "the country limits non-profits."[^13685] The punishment mechanism found previously does not translate to the amount donated—while there is a decrease of `r md_dollar(abs(filter(amount, title == "Human rights | Government")$mu_diff))` in average donations to government-funded human rights NGOs facing crackdown, the chance this difference is less than zero is lower than our threshold (`r filter(amount, title == "Human rights | Government")$p_neq0`).

@@

In contrast, crackdowns have no substantial effect on the amount donated to privately funded humanitarian NGOs (`r filter(amount, title == "Humanitarian assistance | Private")$p_neq0`). No respondent in these conditions mentioned funding sources in their justification, and many explained that they would not give solely because of the crackdown and that they "would keep all \$100 because it would not even go to helping humanitarian efforts due to the new laws."[^8995923] It is possible that donors see government-funded humanitarian NGOs as more legitimate and more capable of handling difficult legal restrictions than privately funded organizations. Donors might assume that an INGO which receives substantial money from USAID would have the support of the U.S. government when facing legal challenges, while an NGO funded by small donors would be less able to confront such challenges. Further research is needed to probe this trend.

@@

Dozens of countries have sought to restrict NGO funding and programming. Between the lack of a local philanthropic culture, an increasingly authoritarian approach that restricts space for domestic philanthropy, and the withdrawal of official governmental aid, many INGOs have turned to foreign private donors as an additional funding source. Since seeking out perfect information about a nonprofit and its activities, especially internationally, is costly and time-consuming, donors use heuristics when deciding whom to support. Existing research has shown that donors take cues and heuristics from social networks, but can other informational cues also influence donor preferences? We argue that domestic political environments of NGO host countries can indeed serve as one such heuristic. Our results show that while crackdowns do not directly influence individual preferences of donation on their own, respondents indicated willingness to donate `r percent_0(filter(amount, title == "Crackdown")$pct_change)` more to legally restricted NGOs. This effect persists when mediated by other heuristics, with respondents giving `r percent_0(filter(amount, title == "Humanitarian assistance")$pct_change)` more to humanitarian NGOs, and `r percent_0(filter(amount, title == "Human rights | Private")$pct_change)` more to privately funded human rights NGOs. Anti-NGO crackdowns can thus serve as a signal of NGO effectiveness.
